WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-27 articles/2014/Article-WSClean/wsclean.tex
Work on article
Final details paper

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements significantly. The source code of our new imager is publicly released.
\end{abstract}

@@

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's $w$-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

@@

TODO: add big full-sky image

\subsection{$w$-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. Computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

@@

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly. Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}. An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean. Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

@@

\subsection{Gridding} \label{sec:gridding}
A gridding convolution kernel improves the accuracy of gridding in $uv$-space. A common kernel function is a windowed sinc function, which acts as a low-pass filter. This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes that arise because of the assumed periodicity during the inverse FFT \citep{post-correlation-filtering}. By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.

@@

\section{Analysis} \label{sec:analysis}
We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in CASA. For the analysis, we use imaging parameters which are common for MWA Imaging. When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used. The number of $w$-projection planes in $w$-projection is kept equal to the number of $w$-layers in $w$-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}. This yields 195 $w$-planes/layers at 10\degree zenith angle. Some configurations with high $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.

@@

In the $\textrm{ZA}=10\degree$ case, CASA is slightly less accurate. The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}. An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analyzed. Its accuracy is higher compared to the normal projection, but its residual noise is also higher. The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.

@@

\begin{figure*}%
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
\caption{}\label{fig:timing-nsamples}%
\end{subfigure}%
\hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
\caption{}\label{fig:timing-zenith-angle}%
\end{subfigure}\\
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
\caption{}\label{fig:timing-resolution}%
\end{subfigure}%
\hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-channels/channels}
\caption{}\label{fig:timing-channels}%
\end{subfigure}\\
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
\caption{}\label{fig:timing-fov}%
\end{subfigure}%
\caption{Imaging performance as a function of several parameters. Error bars show 5$\sigma$ level. Unfinished lines indicate the imager could not run succesfully.}\label{fig:timings}
\end{figure*}

@@

We measured the performance of the imagers using several MWA data sets. Each specific configuration is run 5 times and standard deviations are calculated. The variations between runs is typically a few seconds. In each benchmark, the wall-clock time is measured that is required to produce the synthesized point-spread function and the image itself. No cleaning or prediction was performed. The results are given in Fig.~\ref{fig:timings}.

@@

In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the field of view. The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 10\degree ZA than $w$-projection.

@@

\begin{table*}
\caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree with 1 polarization. Predictions are for $w$-projection with CASA; $w$-stacking with WSClean; $w$-stacked snapshots with WSClean using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using CASA with optimal $\Delta w$. All have approximately equal accuracy.} \label{tbl:computational-cost-predictions}
\begin{tabular}{l|rrrr|rr}
               & \multicolumn{4}{c|}{Predicted computing cost on test computer} & min & best \\
 Configuration & $w$-projection & $w$-stacking & $w$-snapshot & hybrid & computing & FLOPS/float \\
 \hline
 GLEAM    &   67h 7m & 8h 31m &  8h 05m & 15h 18m & 1.2 TFLOPS & $6.9 \times 10^2$    \\
 EMU      &    5.2 d &  2.9 d &   2.9 d & 5.2 d   & 10 TFLOPS & $2.1 \times 10^4$      \\
 MSSS low &  2h 18m  & 0h 59m &  0h 58m & 2h 16m  & 140 GFLOPS & $3.5 \times 10^3$   \\
 MSSS high&  7h 16m  & 3h 52m &  3h 52m & 7h 16m  & 530 GFLOPS & $3.4 \times 10^3$   \\
 LOFAR NL &  50.6 d  &   10 d &   7.2 d & 15.9 d & 33 TFLOPS & $4.5 \times 10^3$ \\
 AARTFAAC &  71h 25m &29h 50m & 29h 50m & 51h 34m & 4.1 TFLOPS & $6.8 \times 10^2$   \\
 VLSS     &  0h 12m  & 0h 12m &  0h 02m &  0h 12m & 4.1 GFLOPS & $12.1 \times 10^3$     \\
 MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d & 21 TFLOPS & $6.8 \times 10^2$ \\
% SKA1 AA inner&392 d &  228 d &   228 d &   392 d & & $6.8 \times 10^2$   \\
 SKA1 AA core&1581 d &  911 d &   911 d &  1571 d &  3 PFLOPS & $6.8 \times 10^2$ \\
 SKA1 AA full& 633 yr& 48.8 yr & 48.8 yr& 84.2 yr & 59 PFLOPS & $6.8 \times 10^2$ \\
\end{tabular}
\end{table*}

\subsection{Combination of $w$-projection and $w$-stacking}
To combine the $w$-projection and $w$-stacking algorithms, small $w$-corrections are made before the FFTs using a $w$-correcting kernel and large $w$-terms are corrected after the FFTs by gridding onto several $w$-layers. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_{\max}$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_{\max}}{\Delta w} t_\textrm{CASA}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_{\max}}$ and $w'_{\max} = \Delta w$. For MWA observations, the optimal value for $\Delta w$ is about unity, and improves the speed of $w$-projection with about a factor of four. However, the performance of the hybrid method is still about a factor of two lower than our pure-stacking implementation. This could again be the difference in optimisation choices between CASA and WSClean, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.

@@

\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes. We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}. The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}. In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams. The image size is set to 5 times the FOV width divided by the resolution. The factor of five is to allow better cleaning accuracy and source-position fitting, as is common. The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).

The results are summarized in Table~\ref{tbl:computational-cost-predictions}. WSClean is in most situations predicted to be 2--3 times faster than CASA. The MWA is somewhat of an exception, where WSClean is 8$\times$ faster. The main cause for this is the wide field of view of the MWA. The $w$-snapshot method does not improve much over normal WSClean operation in any of the cases, except the VLSS case. We think this is because the VLSS had a modest data rates compared to the other configurations, making the constant cost per $w$-plane/layer stand out. The $w$-snapshot might become more valuable at higher zenith angles (TODO do I want to predict that too?) The hybrid method is in all situations appromixately a factor of two slower than WSClean.

Imaging the full field of view with the full SKA becomes very expensive due to the resolution of $18\textrm{k} \times 18$k pixels. This translates to a computing power of $\sim60$ PetaFLOPS.

\section{Conclusions} \label{sec:conclusions}
We have shown that the $w$-stacking algorithm is well suited for imaging MWA observations. The WSClean $w$-stacking implementation is faster than CASA's $w$-projection algorithm in all common MWA imaging configurations, leading up to an order of magnitude difference already at a relatively small zenith angle of $10$\degree. It is also slightly more accurate. For approximately zenith angles $>15\degree$ or field-of-view diameters $>35\degree$, snapshot imaging becomes faster. This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used. Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots is worthwhile for the MWA only at very low elevations or large fields of view. Extrapolation of the computing cost predicts that this holds for most arrays, including SKA low. A hybrid between $w$-stacking and $w$-projection seems not to improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.

The available SKA computing power is estimated to be around $\sim$100 PetaFLOPS. Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone. It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm. Clearly it remains challenging to perform widefield imaging with acceptable performance, but some optimisations could be made for the SKA. For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging. Moreover, although longer baselines make the imaging more expensive, it is likely that it is not often required to image the full field of view when using the longest baselines.

@@

>>>>

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. The performance dependencies of CASA's $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers. On MWA data, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.  We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration. Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques. We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's $w$-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

@@

TODO: add big full-sky image (PKS J2358-6054 is visible)

\subsection{$w$-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. Imaging computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

@@

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly. Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.

An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean. The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibrate on Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) image each of the seven 112 s intervals separately. Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

@@

\subsection{Gridding} \label{sec:gridding}
A gridding convolution kernel improves the accuracy of gridding in $uv$-space. A common kernel function is a windowed sinc function, which acts as a low-pass filter. This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes \citep{post-correlation-filtering}. By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.

@@

\subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
Typically, for cleaning it is desired that images have a pixel size at least 5 times smaller than the synthesized-beam width. This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions. This factor of $\sim$5 is normally taken into account in the overal imaging resolution, i.e., the image size is increased during inversion. For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels. It will affect the inverse FFT, but the relative cost of this step is negligable in the total cost. The $w$-stacking method is however significant affected by the image resolution, because it performs many inverse FFTs.

The spatial frequencies in the output image are band limited by the synthesized beam. Therefore, it is not required to perform the inversion at higher resolution. As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution. After an upscaled image has been cleaned, a model is created at the same (high) resolution. Because the model consists of delta functions, the spatial frequencies in the model are not band limited. Therefore, downscaling the model will remove information from the model. However, only the low spatial frequencies of this image will be used in the prediction. As long as downscaling does not modify the low-frequency components, the output of the prediction step will not change. This requires low-pass filtering the model before decimation.

We have implemented an option in WSClean to automatically decrease the inversion resolution to the Nyquist limit,
\begin{equation}\label{eq:nyquist-resolution}
 N_\textrm{Nyquist} = 2 \frac{N_\textrm{pix} S_\textrm{pix}}{S_\textrm{synth beam}},
\end{equation}
where $N_\textrm{Nyquist}$ the resolution in pixels used during inversion, $N_\textrm{pix}$ the requested number of pixels of the image along one side, $S_\textrm{pix}$ the requested angular pixel scale and $S_\textrm{synth beam}$ the minimum angular size of the synthesized beam. Before cleaning, the low-resolution image is interpolated with a procedure consisting of i) a low-resolution FFT; ii) zero padding; and iii) a high-resolution inverse FFT. Such interpolation assumes the image to be periodic, but this is already assumed during the inversion process. After cleaning, the model is decimated by the inverse procedure, consisting of i) a high-resolution FFT; ii) truncation; and iii) a low-resolution inverse FFT. With common settings, this procedure decreases the inversion imaging resolution with a factor of 2.5. Without further correction, such a decrease would lower the positional accuracy with which visibilities are gridded onto the $uv$-plane. This can be corrected by increasing the oversampling rate, which has almost no affect on performance.

Recreating Fig.~\ref{fig:vela-projection-example} using both the snapshot method and this optimization lowers the computational cost from 41~min to 24~min. Of the 24~min, 77\% is spent on cleaning approximately 100,000 components. The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy. We think this can be explained by the non-linear behaviour of clean.

\section{Analysis} \label{sec:analysis}
We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in CASA. For the analysis, we use imaging parameters which are common for MWA Imaging. When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used. The number of $w$-projection planes in $w$-projection is kept equal to the number of $w$-layers in $w$-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}. This yields 195 $w$-planes/layers at 10\degree zenith angle. Several configurations with large $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.

@@

In the $\textrm{ZA}=10\degree$ case, CASA is slightly less accurate. The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}. An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analyzed. Its accuracy is higher compared to the normal projection, but its residual noise is also higher. The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy. We have performed tests with and without the optimization of \S\ref{sec:decreasing-inversion-resolution}. They yield identical numbers.

@@

\begin{figure*}%
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
\caption{}\label{fig:timing-nsamples}%
\end{subfigure}%
\hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
\caption{}\label{fig:timing-zenith-angle}%
\end{subfigure}\\
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
\caption{}\label{fig:timing-resolution}%
\end{subfigure}%
\hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-channels/channels}
\caption{}\label{fig:timing-channels}%
\end{subfigure}\\
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
\caption{}\label{fig:timing-fov}%
\end{subfigure}%
\caption{Imaging performance as a function of several parameters. Error bars show 5$\sigma$ level. Unfinished lines indicate the imager could not run the specific configuration succesfully.}\label{fig:timings}
\end{figure*}

@@

We measured the performance of the imagers using several MWA data sets. Each specific configuration is run 5 times and standard deviations are calculated. The variations between runs is typically a few seconds. In each benchmark, the wall-clock time is measured that is required to produce the synthesized point-spread function and the image itself. No cleaning or prediction was performed, and the optimization of \S\ref{sec:decreasing-inversion-resolution} is not used. The results are given in Fig.~\ref{fig:timings}.

@@

In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the field of view. The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 0\degree ZA than $w$-projection. Imaging the primary MWA beam requires approximately a resolution of 3k for cleaning. If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1.5k. At ZA=10\degree, this saves about a factor of 2 in computing cost. The benefit of the optimization increases with resolution. At 10k resolution, cost is decreased by an order of magnitude.

@@

\begin{table*}
\caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree with 1 polarization. Predictions are for $w$-projection with CASA; $w$-stacking with WSClean; $w$-stacked snapshots with WSClean using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using CASA with optimal $\Delta w$. All have approximately equal accuracy.} \label{tbl:computational-cost-predictions}
\begin{tabular}{l|rrrr|rr}
               & \multicolumn{4}{c|}{Predicted computing cost on test computer} & min & best \\
 Configuration & $w$-projection & $w$-stacking & $w$-snapshot & hybrid & computing & FLOPS/float \\
 \hline
 GLEAM    &  65h 25m & 8h 04m &  8h 03m & 14h 10m   & 1.1 TFLOPS & $6.9 \times 10^2$    \\
 EMU      &    5.2 d &  2.9 d &   2.9 d & 4.4 d     & 9.7 TFLOPS & $2.1 \times 10^4$      \\
 MSSS low &  2h 16m  & 0h 57m &  0h 57m & 2h 10m    & 130 GFLOPS & $3.5 \times 10^3$   \\
 MSSS high&  7h 16m  & 3h 52m &  3h 52m & 5h 44m    & 530 GFLOPS & $3.4 \times 10^3$   \\
 LOFAR NL &  50.2 d  &  7.3 d &   7.2 d & 12.9 d    &  24 TFLOPS & $4.5 \times 10^3$ \\
 AARTFAAC &  2.5 d   &  1.2 d &   1.2 d &   2.5 d   & 4.1 TFLOPS & $6.8 \times 10^2$   \\
 VLSS     &  0h 09m  & 0h 01m &  0h 01m &  0h 08m   & 2.2 GFLOPS & $12.1 \times 10^3$     \\
 MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d   &  21 TFLOPS & $6.8 \times 10^2$ \\
% SKA1 AA inner&392 d &  228 d &   228 d &   392 d & & $6.8 \times 10^2$   \\
 SKA1 AA core&1581 d &  911 d &   911 d &  1570.8   & 3.0 PFLOPS & $6.8 \times 10^2$ \\
 SKA1 AA full& 643 yr& 48.8 yr & 48.8 yr& 84.1 yr   &  59 PFLOPS & $6.8 \times 10^2$ \\
\end{tabular}
\end{table*}

\subsection{Combination of $w$-projection and $w$-stacking}
To combine the $w$-projection and $w$-stacking algorithms, small $w$-corrections are made before the FFTs using a $w$-correcting kernel and large $w$-terms are corrected after the FFTs by gridding onto several $w$-layers. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_{\max}$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_{\max}}{\Delta w} t_\textrm{CASA}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_{\max}}$ and $w'_{\max} = \Delta w$. For MWA observations, the optimal value for $\Delta w$ is about unity, and improves the speed of $w$-projection with about a factor of four. However, the performance of the hybrid method is still about a factor of two lower than our pure-stacking implementation. This could again be the difference in optimisation choices between CASA and WSClean, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile. All configurations except VLSS and GLEAM have optimal values for $\Delta w$ much smaller than one, suggesting all $w$-corrections should be done by stacking instead of projection.

@@

\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes. We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}. The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}. In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams. The image size is set to 2 times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}. The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).

The results are summarized in Table~\ref{tbl:computational-cost-predictions}. WSClean is in most situations predicted to be 2--3 times faster than CASA. WSClean has the largest benefit on full LOFAR, MWA and the full SKA, where WSClean is respectively 7, 8$\times$ and 12 times faster. The $w$-snapshot method does not improve much over normal WSClean operation in any of the cases, except the VLSS case. We think this is because the VLSS had a modest data rates compared to the other configurations, making the constant cost per $w$-plane/layer stand out. The $w$-snapshot might become more valuable at higher zenith angles. The hybrid method is in all situations appromixately a factor of two slower than WSClean.

Imaging the full field of view with the full SKA becomes very expensive due to the high frequency and time resolution, and image resolution of $7.2\textrm{k} \times 7.2$k pixels. This translates to a computing power requirement of $\sim60$ PetaFLOPS.

\section{Conclusions} \label{sec:conclusions}
We have shown that the $w$-stacking algorithm is well suited for imaging MWA observations. The WSClean $w$-stacking implementation is faster than CASA's $w$-projection algorithm in all common MWA imaging configurations, leading up to an order of magnitude difference already at a relatively small zenith angle of $10$\degree. It is also slightly more accurate. For approximately zenith angles $>15\degree$ or field-of-view diameters $>35\degree$, snapshot imaging becomes faster. This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used. Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots is worthwhile for the MWA only at very low elevations or large fields of view. Extrapolation of the computing cost predicts that this holds for most arrays, including SKA low. A hybrid between $w$-stacking and $w$-projection seems not to improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly. Optimizing image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticable effect on the accuracy.

The available SKA computing power is estimated to be around $\sim$100 PetaFLOPS. Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone. It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm. Clearly it remains challenging to perform widefield imaging with acceptable performance, but some optimisations could be made for the SKA. For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging. Moreover, although longer baselines make the imaging more expensive, it is likely that it is not always required to image the full field of view when using the longest baselines.

@@

WSClean is currently not able to perform multi-scale clean. Results of applying CASA's multi-scale clean on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive. Imaging a one-minute observation takes 30 hours of computational time without $w$-projection. We plan to implement some form of multi-scale cleaning in WSClean. It is likely that additional optimizations need to be made to be able to do this with acceptable performance.
